---
title: "Getting Started with OmicSelector 2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with OmicSelector 2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

**OmicSelector 2.0** is a PhD-level toolkit for high-dimensional biomarker discovery that guarantees scientific validity through rigorous machine learning methodology.

## Core Philosophy

> "Optimization without validation is hallucination."

OmicSelector 2.0 prioritizes:

1. **Zero Data Leakage** - All preprocessing occurs within CV folds
2. **Feature Stability** - Nogueira Stability Index over raw accuracy
3. **Reproducibility** - renv lockfiles, Docker containers, deterministic pipelines

# Quick Start (5 minutes)

## Installation

```{r install}
# From GitHub
remotes::install_github("kstawiski/OmicSelector")

# Load the package
library(OmicSelector)
```

## Basic Workflow

### Step 1: Create Pipeline

```{r create-pipeline}
# Load your data
data <- read.csv("expression.csv")

# Create pipeline
pipeline <- OmicPipeline$new(
  data = data,
  target = "outcome",
  positive = "Case"
)
```

### Step 2: Build Graph Learner

The GraphLearner encapsulates all preprocessing within CV folds:

```{r create-learner}
# Impute → Scale → Filter → Model (all inside CV)
learner <- pipeline$create_graph_learner(
  filter = "anova",       # Feature selection method
  model = "ranger",       # Random Forest
  n_features = 20,        # Select top 20 features
  scale = TRUE            # Standardize features
)
```

### Step 3: Run Nested Cross-Validation

```{r run-benchmark}
# Create benchmark service with nested CV
service <- BenchmarkService$new(
  task = pipeline,
  outer_folds = 5,   # Evaluation folds
  inner_folds = 3    # Feature selection folds
)

# Add learner and run
service$add_learner(learner)
result <- service$run()
```

### Step 4: Check Stability

```{r check-stability}
# Compute feature selection stability
stability <- compute_stability_from_resample(
  result$resample_result,
  all_features = pipeline$get_feature_names()
)

print(stability)
# Nogueira Stability Index: 0.78
# Interpretation: Good - Reasonably stable feature selection
```

### Step 5: Generate Report

```{r generate-report}
# Create TRIPOD+AI compliant report
generate_tripod_report(
  results = result,
  output_file = "my_analysis_report.html"
)
```

# Multi-Omics Analysis

OmicSelector 2.0 supports multi-omics data with automatic feature namespacing:

```{r multi-omics}
# Prepare multi-omics data as named list
multi_data <- list(
  rna = rna_expression,      # Gene expression
  mirna = mirna_expression,  # miRNA expression
  prot = protein_levels      # Proteomics
)

# Target column should be in ONE modality
multi_data$rna$outcome <- clinical_outcome

# Create pipeline - features are namespaced (rna::gene1, mirna::mir21)
pipeline <- OmicPipeline$new(
  data = multi_data,
  target = "outcome",
  positive = "Case"
)

# Check modalities
pipeline$get_modality_info()
#>   modality n_features n_samples has_target
#> 1      rna       5000       200       TRUE
#> 2    mirna        800       200      FALSE
#> 3     prot       1500       200      FALSE
```

# Parallel Processing

Enable parallelization for faster benchmarking:
```{r parallel}
# Setup parallel processing
setup_parallel(workers = 4)

# Run benchmark (uses all workers)
result <- service$run(parallel = TRUE)

# Reset to sequential
reset_parallel()
```

# Command-Line Interface

OmicSelector provides a CLI for batch processing:

```bash
# Run benchmark from config file
omicselector run --config=config.yaml --parallel=4 --seed=42

# Generate report from results
omicselector report --results=results/benchmark.rds --format=html

# Validate data before analysis
omicselector validate --data=expression.csv --target=outcome
```

# Key Concepts

## Why Nested Cross-Validation?

Standard CV leaks information when feature selection happens before the split:

```
WRONG (Leaky):
1. Select features on ALL data  ← Leakage!
2. Split into train/test
3. Train model
4. Evaluate (overly optimistic)

RIGHT (OmicSelector):
1. Split into outer folds
   For each outer fold:
   2. Split training into inner folds
      For each inner fold:
      3. Select features on inner training only
      4. Train model
   5. Aggregate best features
   6. Final model on outer training
   7. Evaluate on outer test (unbiased)
```

## Why Stability Matters

High accuracy with unstable features = overfitting:

| Stability Index | Interpretation |
|-----------------|----------------|
| ≥ 0.9 | Excellent - Very stable |
| 0.7 - 0.9 | Good - Reasonably stable |
| 0.5 - 0.7 | Moderate - Some instability |
| 0.3 - 0.5 | Poor - Likely overfitting |
| < 0.3 | Very Poor - Random/overfit |

# Next Steps

- **Phase 5 Features**: Try GOF filters, AutoXAI, and stability ensembles
  - See the [Phase 5 Advanced Features](phase5-advanced.html) vignette
- **Advanced Usage**: Explore custom filter combinations
- **External Validation**: Use frozen preprocessing for new data
- **TRIPOD+AI Compliance**: Review generated reports for publication

# Phase 5 Quick Preview

```{r phase5-preview}
# GOF filters for sparse/zero-inflated data
learner <- pipeline$create_graph_learner(
 filter = "gof_ks",    # Kolmogorov-Smirnov filter
  model = "ranger",
  n_features = 30
)

# AutoXAI with correlation warnings
xai <- xai_pipeline(learner, task, cor_threshold = 0.7)
plot_xai_importance(xai)

# Bootstrap stability ensemble
ensemble <- create_stability_ensemble(preset = "default", n_bootstrap = 100)
ensemble$fit(task, seed = 42)
stable_features <- ensemble$get_feature_importance(30)

# Bayesian hyperparameter tuning
autotuner <- make_autotuner_glmnet(task, n_evals = 20)
autotuner$train(task)
```

# Getting Help

- GitHub Issues: https://github.com/kstawiski/OmicSelector/issues
- Documentation: https://biostat.umed.pl/OmicSelector/
