---
title: "Modern OmicSelector Workflow: Tidymodels and TRIPOD+AI Compliance"
author: "OmicSelector Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Modern OmicSelector Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Set to TRUE when ready to build
)
```

## Introduction

This vignette demonstrates the modern OmicSelector workflow, which includes:

1. **Tidymodels integration** for modern machine learning
2. **Nested cross-validation** to prevent data leakage
3. **TRIPOD+AI compliance** for transparent reporting
4. **PROBAST+AI assessment** for risk of bias evaluation
5. **Clinical utility metrics** for practical applicability

The modernized OmicSelector maintains backward compatibility with existing caret-based workflows while providing access to state-of-the-art methods.

## Why Modern OmicSelector?

### Problems with Traditional Approaches

Traditional biomarker discovery often suffers from:

- **Data leakage**: Preprocessing on full dataset before cross-validation
- **Optimistic bias**: Feature selection outside of resampling loops
- **Poor reporting**: Missing key details for reproducibility
- **Lack of clinical utility assessment**: Focus on AUC without calibration or decision curves

### OmicSelector 2.0 Solutions

The modernized OmicSelector addresses these issues through:

- **Rigorous nested CV**: All preprocessing inside resampling folds
- **Automated compliance**: TRIPOD+AI and PROBAST+AI reporting
- **Clinical metrics**: Calibration, decision curves, net benefit
- **Modern ML framework**: Tidymodels with extensive algorithm support

## Installation

```{r install, eval=FALSE}
# Install development version from GitHub
remotes::install_github("kstawiski/OmicSelector")

# Load required packages
library(OmicSelector)
library(tidymodels)
library(tidyverse)
```

## Basic Workflow

### 1. Load and Prepare Data

```{r data}
library(OmicSelector)

# For this example, we'll create synthetic biomarker data
# In practice, you would load your own omics data
set.seed(2024)

n_samples <- 150
n_features <- 100

# Simulate gene expression data
expression_data <- matrix(
  rnorm(n_samples * n_features, mean = 8, sd = 2),
  nrow = n_samples,
  ncol = n_features
)
colnames(expression_data) <- paste0("gene_", 1:n_features)

# Add some truly predictive features
expression_data[1:75, 1:5] <- expression_data[1:75, 1:5] + 2
expression_data[76:150, 1:5] <- expression_data[76:150, 1:5] - 2

# Create outcome
outcome <- factor(c(rep("case", 75), rep("control", 75)))

# Combine into data frame
data <- as.data.frame(expression_data)
data$diagnosis <- outcome

# View data structure
dim(data)
table(data$diagnosis)
```

### 2. Define Preprocessing Recipe

One of the key advantages of tidymodels is the `recipes` framework for preprocessing:

```{r recipe}
library(recipes)

# Define preprocessing steps
# CRITICAL: This will be applied INSIDE each CV fold
preprocessing_recipe <- recipe(diagnosis ~ ., data = data) %>%
  # Normalize all numeric predictors
  step_normalize(all_numeric_predictors()) %>%
  # Remove zero-variance features
  step_zv(all_predictors()) %>%
  # Remove highly correlated features (optional)
  step_corr(all_numeric_predictors(), threshold = 0.9)

# Preview the recipe (not yet trained)
preprocessing_recipe
```

### 3. Define Model Specifications

Specify multiple models to compare:

```{r models}
library(parsnip)

# Random Forest
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# XGBoost
xgb_spec <- boost_tree(
  mtry = tune(),
  trees = tune(),
  min_n = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Elastic Net (for comparison - simpler model)
glmnet_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet")

# Combine into list
models <- list(
  random_forest = rf_spec,
  xgboost = xgb_spec,
  elastic_net = glmnet_spec
)
```

### 4. Run Nested Cross-Validation

This is the core of the modern OmicSelector approach:

```{r nested_cv}
# Run nested CV with proper validation
result <- OmicSelector_nested_cv(
  data = data,
  outcome = "diagnosis",
  outer_folds = 5,
  inner_folds = 5,
  preprocessing_recipe = preprocessing_recipe,
  feature_selection_method = "stability_selection",
  feature_selection_params = list(
    n_iterations = 100,
    selection_threshold = 0.6
  ),
  models = models,
  calibrate = TRUE,
  seed = 2024
)

# View results
print(result)
summary(result)
```

The nested CV procedure ensures:

1. **Outer loop** (5 folds): Provides unbiased performance estimates
2. **Inner loop** (5 folds): Used for hyperparameter tuning
3. **Preprocessing** happens independently in each fold
4. **Feature selection** happens independently in each fold
5. **No data leakage** between train and test sets

### 5. Examine Results

```{r examine}
# View performance across models
result$aggregated_results

# View selected features (may vary by fold)
result$selected_features

# View calibration results
result$calibration

# Get best model
best_model_name <- names(which.max(
  sapply(result$aggregated_results, function(x) x$mean_performance$roc_auc)
))
print(paste("Best model:", best_model_name))
```

## TRIPOD+AI Compliance

Generate a comprehensive TRIPOD+AI report:

```{r tripod}
# Define study information
study_info <- list(
  title = "Discovery of Gene Expression Biomarkers for Disease X",
  authors = c("Smith J", "Jones A", "Brown B"),
  study_design = "cross-sectional",
  data_source = "Gene Expression Omnibus (GSE12345)",
  inclusion_criteria = "Patients diagnosed with disease X between 2020-2023",
  exclusion_criteria = "Patients with confounding conditions",
  outcome_definition = "Binary disease status confirmed by clinical diagnosis",
  candidate_predictors = "Genome-wide gene expression (RNA-seq)",
  sample_size_justification = "Based on pilot study power analysis (n=150)",
  intended_use = "To identify a gene signature for risk stratification"
)

# Generate TRIPOD+AI report
tripod_report <- OmicSelector_tripod_report(
  model_result = result,
  study_info = study_info,
  output_format = "html",
  output_file = "tripod_report.html",
  include_checklist = TRUE,
  include_recommendations = TRUE
)

# View report
print(tripod_report)

# Check completeness
tripod_report$completeness_score
```

The TRIPOD+AI report includes all 27 items required for transparent reporting:

- Title and abstract
- Introduction (rationale, objectives)
- Methods (data sources, participants, outcomes, predictors, analysis)
- Results (performance, calibration, model specification)
- Discussion (interpretation, limitations)

## PROBAST+AI Risk of Bias Assessment

Assess potential sources of bias:

```{r probast}
# Generate PROBAST assessment
probast <- OmicSelector_probast(
  model_result = result,
  assessment = list(
    # You can provide manual assessments for domains
    # that can't be automatically determined
    participants = list(
      risk = "Low",
      note = "Representative sample from tertiary care center"
    ),
    predictors = list(
      risk = "Low",
      note = "RNA-seq performed using standard protocols"
    ),
    outcome = list(
      risk = "Low",
      note = "Outcome determined by expert clinicians"
    )
    # Analysis domain will be automatically assessed as "Low"
    # because nested CV was used
  ),
  output_format = "html",
  output_file = "probast_assessment.html"
)

# View assessment
print(probast)

# Overall risk
probast$overall_risk

# Domain-specific risks
probast$domain_assessments

# Recommendations
probast$recommendations
```

## Advanced Features

### Feature Stability Analysis

Examine which features are consistently selected across folds:

```{r stability}
# Extract selected features from all folds
all_features <- unlist(result$selected_features)

# Calculate selection frequency
feature_freq <- sort(table(all_features), decreasing = TRUE)

# Plot top 20 most stable features
library(ggplot2)

feature_stability <- data.frame(
  feature = names(head(feature_freq, 20)),
  frequency = as.numeric(head(feature_freq, 20)) / result$metadata$outer_folds
)

ggplot(feature_stability, aes(x = reorder(feature, frequency), y = frequency)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Feature Selection Stability",
    x = "Feature",
    y = "Selection Frequency (across folds)"
  ) +
  theme_minimal()
```

### Model Calibration

Assess whether predicted probabilities match observed frequencies:

```{r calibration}
# Calibration is automatically assessed if calibrate = TRUE
# View calibration metrics
result$calibration

# You can also use the dedicated calibration function
# for more detailed analysis
predictions <- result$predictions  # Out-of-fold predictions

OmicSelector_calibrate(
  predictions = predictions,
  outcomes = data$diagnosis,
  method = "isotonic",
  plot = TRUE
)
```

### Decision Curve Analysis

Assess clinical utility across decision thresholds:

```{r decision_curve}
# Decision curve analysis
OmicSelector_decision_curve(
  model = result,
  data = data,
  outcome = "diagnosis",
  thresholds = seq(0.01, 0.99, by = 0.01),
  plot = TRUE
)

# Clinical impact curve
OmicSelector_clinical_impact(
  model = result,
  data = data,
  outcome = "diagnosis",
  population_size = 1000
)
```

## Simple Fit (Non-Nested)

For quick model fitting (e.g., during exploration), use `OmicSelector_fit`:

```{r simple_fit}
# Quick fit with automatic framework selection
simple_result <- OmicSelector_fit(
  data = data,
  outcome = "diagnosis",
  method = "auto",
  algorithm = "ranger",
  preprocessing = list(
    normalize = TRUE,
    remove_zero_variance = TRUE
  ),
  resampling = list(method = "cv", folds = 5),
  seed = 2024
)

# View results
print(simple_result)
```

**Note**: `OmicSelector_fit` is faster but provides less rigorous validation than nested CV. Use it for exploration, but always use nested CV for final results.

## Backward Compatibility

### Using Caret

The modern framework maintains compatibility with caret:

```{r caret}
# Force caret backend
caret_result <- OmicSelector_fit(
  data = data,
  outcome = "diagnosis",
  method = "caret",  # Explicitly use caret
  algorithm = "rf",
  resampling = list(method = "cv", folds = 5),
  seed = 2024
)

print(caret_result)
```

### Migrating Old Code

Use the migration helper to convert old code:

```{r migrate, eval=FALSE}
# Not yet implemented - coming soon
old_code <- '
train_control <- trainControl(method = "cv", number = 5)
model <- train(diagnosis ~ ., data = data, method = "rf", trControl = train_control)
'

# Convert to tidymodels
new_code <- OmicSelector_migrate(old_code)
```

## Model Card and Documentation

Create a comprehensive model card for sharing:

```{r model_card}
model_card <- OmicSelector_model_card(
  model_result = result,
  model_details = list(
    name = "Disease X Gene Expression Classifier",
    version = "1.0",
    date = Sys.Date(),
    authors = "Research Team",
    license = "CC-BY-4.0"
  ),
  intended_use = "
    This model is intended for research purposes to identify patients at high
    risk of disease X based on gene expression profiles. It should be used in
    conjunction with clinical judgment and is not approved for clinical use.
  ",
  limitations = "
    - Developed on samples from a single center
    - Limited to adult patients
    - Requires fresh RNA samples
    - Performance may vary in different populations
  ",
  ethical_considerations = "
    Patient consent was obtained for all samples. Data privacy was maintained
    throughout the study. The model should not be used to deny care or discriminate.
  ",
  output_file = "model_card.json"
)

print(model_card)
```

## Best Practices

### 1. Always Use Nested CV for Final Results

```{r best_practice_1}
# ✓ GOOD: Nested CV for unbiased estimates
result_good <- OmicSelector_nested_cv(
  data = data,
  outcome = "diagnosis",
  outer_folds = 5,
  inner_folds = 5,
  models = models
)

# ✗ BAD: Single CV can be optimistically biased
# (okay for exploration, not for final reporting)
result_exploratory <- OmicSelector_fit(
  data = data,
  outcome = "diagnosis",
  method = "tidymodels",
  resampling = list(method = "cv", folds = 5)
)
```

### 2. Preprocess Inside Resampling Folds

```{r best_practice_2}
# ✓ GOOD: Recipe applied inside CV folds
recipe <- recipe(diagnosis ~ ., data = data) %>%
  step_normalize(all_numeric_predictors())

result <- OmicSelector_nested_cv(
  data = data,
  outcome = "diagnosis",
  preprocessing_recipe = recipe,
  models = models
)

# ✗ BAD: Preprocessing before CV causes leakage
# data_normalized <- scale(data[, -ncol(data)])  # Don't do this!
# result <- OmicSelector_nested_cv(data = data_normalized, ...)
```

### 3. Report Complete TRIPOD+AI

```{r best_practice_3}
# Always generate and include TRIPOD+AI report
tripod <- OmicSelector_tripod_report(
  model_result = result,
  study_info = complete_study_info,  # Provide all details
  output_file = "supplementary_tripod.html"
)

# Aim for >90% completeness
stopifnot(tripod$completeness_score$percentage >= 90)
```

### 4. Assess Calibration

```{r best_practice_4}
# Always check calibration, not just discrimination (AUC)
result <- OmicSelector_nested_cv(
  data = data,
  outcome = "diagnosis",
  models = models,
  calibrate = TRUE  # Enable calibration assessment
)

# Review calibration metrics
result$calibration
```

## Troubleshooting

### Memory Issues

```{r troubleshooting_memory}
# For large datasets, consider:

# 1. Reduce number of outer folds
result <- OmicSelector_nested_cv(
  data = large_data,
  outcome = "diagnosis",
  outer_folds = 3,  # Instead of 5
  inner_folds = 3,
  models = models,
  save_predictions = FALSE  # Don't store all predictions
)

# 2. Use simpler models
simple_models <- list(
  elastic_net = glmnet_spec  # Faster than random forest
)

# 3. Reduce feature set first
top_features <- OmicSelector_feature_selection(
  data = large_data,
  outcome = "diagnosis",
  method = "variance",
  n_features = 500  # Keep top 500
)
```

### Parallel Processing

```{r troubleshooting_parallel}
# Enable parallel processing for faster computation
library(doParallel)

# Set up parallel backend
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

# Run nested CV with parallel processing
result <- OmicSelector_nested_cv(
  data = data,
  outcome = "diagnosis",
  models = models,
  parallel = TRUE
)

# Clean up
stopCluster(cl)
```

## Next Steps

### Multi-omics Integration

See the multi-omics vignette for DIABLO and MOFA integration:

```{r next_multiomics, eval=FALSE}
vignette("multiomics_integration", package = "OmicSelector")
```

### Survival Analysis

For time-to-event outcomes:

```{r next_survival, eval=FALSE}
vignette("survival_analysis", package = "OmicSelector")
```

### Advanced Feature Selection

For stability selection and Model-X knockoffs:

```{r next_features, eval=FALSE}
vignette("advanced_feature_selection", package = "OmicSelector")
```

## References

1. Collins GS, Moons KGM, Dhiman P, et al. TRIPOD+AI statement: updated guidance for reporting clinical prediction models that use regression or machine learning methods. BMJ 2024;385:e078378.

2. Wolff RF, Moons KGM, Riley RD, et al. PROBAST: A Tool to Assess the Risk of Bias and Applicability of Prediction Model Studies. Ann Intern Med. 2019;170(1):51-58.

3. Kuhn M, Silge J. Tidy Modeling with R. https://www.tmwr.org/

4. Nogueira S, Sechidis K, Brown G. On the Stability of Feature Selection Algorithms. Journal of Machine Learning Research. 2017;18(174):1-54.

## Session Info

```{r session_info}
sessionInfo()
```
