---
title: "Modern OmicSelector Workflow with Tidymodels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modern OmicSelector Workflow with Tidymodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE once we have example data
)
```

# Introduction

OmicSelector 2.0 introduces a modern workflow based on the tidymodels framework, with rigorous nested cross-validation to prevent data leakage and TRIPOD+AI compliance for clinical prediction models.

This vignette demonstrates the key features of the modernized package.

## Installation

```{r setup}
library(OmicSelector)
library(tidymodels)
library(dplyr)
library(ggplot2)
```

# Problem: Data Leakage in Traditional Workflows

Data leakage occurs when information from the test set "leaks" into the training process, leading to optimistically biased performance estimates.

## Common Leakage Sources

1. **Preprocessing before splitting**: Normalizing or scaling the entire dataset
2. **Feature selection on full data**: Selecting features before train/test split
3. **Hyperparameter tuning with test data**: Using test set for any decisions

## OmicSelector 2.0 Solution

All preprocessing, feature selection, and hyperparameter tuning happen **inside** resampling folds.

# Basic Workflow: Nested Cross-Validation

## Step 1: Load Your Data

```{r}
# Example with built-in data
data(miR_Asakura)  # Replace with your data

# Check data structure
str(miR_Asakura)
```

## Step 2: Define Preprocessing Recipe

```{r}
# Create a recipe for preprocessing
# This will be applied INSIDE each fold
library(recipes)

preprocessing_recipe <- recipe(Class ~ ., data = miR_Asakura) %>%
  # Remove zero-variance features
  step_zv(all_predictors()) %>%
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors()) %>%
  # Handle missing values if needed
  step_impute_median(all_numeric_predictors())

# Preview the recipe (but don't prep yet!)
preprocessing_recipe
```

## Step 3: Define Models to Compare

```{r}
library(parsnip)

# Define multiple models with hyperparameters to tune
models <- list(
  # Random Forest
  ranger = rand_forest(
    mtry = tune(),           # Number of variables to try at each split
    min_n = tune(),          # Minimum node size
    trees = 1000
  ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("classification"),

  # XGBoost
  xgboost = boost_tree(
    mtry = tune(),
    trees = tune(),
    min_n = tune(),
    tree_depth = tune(),
    learn_rate = tune()
  ) %>%
    set_engine("xgboost") %>%
    set_mode("classification"),

  # Elastic Net (LASSO + Ridge)
  glmnet = logistic_reg(
    penalty = tune(),
    mixture = tune()
  ) %>%
    set_engine("glmnet") %>%
    set_mode("classification")
)
```

## Step 4: Run Nested Cross-Validation

```{r}
# Set seed for reproducibility
set.seed(123)

# Run nested CV with feature selection
result <- OmicSelector_nested_cv(
  data = miR_Asakura,
  outcome = "Class",

  # Cross-validation strategy
  outer_folds = 5,    # For performance estimation
  inner_folds = 5,    # For hyperparameter tuning

  # Preprocessing (applied inside folds)
  preprocessing_recipe = preprocessing_recipe,

  # Feature selection (applied inside folds)
  feature_selection_method = "boruta",
  max_features = 20,

  # Models to compare
  models = models,

  # Number of hyperparameter combinations to try
  tune_grid = 10,

  # Assess calibration
  calibrate = TRUE,

  # Use parallel processing
  parallel = TRUE,
  cores = 4
)
```

## Step 5: View Results

```{r}
# Print summary
print(result)

# Detailed summary
summary(result)

# Visualize results
plot(result)
```

## Step 6: Examine Feature Stability

```{r}
# Which features were selected consistently?
result$feature_stability

# Most stable features
head(result$feature_stability$stable_features, 20)

# Jaccard similarity between folds
result$feature_stability$mean_jaccard_similarity
```

# TRIPOD+AI Compliance Reporting

Clinical prediction models should follow TRIPOD+AI guidelines for transparent reporting.

## Generate TRIPOD+AI Report

```{r}
# Define study information
study_info <- list(
  title = "MicroRNA-Based Diagnostic Model for Cancer Detection",
  authors = c("Smith J", "Jones A", "Brown B"),
  objective = "To develop and internally validate a diagnostic model for cancer detection using microRNA expression profiles",
  data_source = "RNA-seq data from tumor and matched normal tissues collected at Academic Medical Center between 2018-2022",
  eligibility_criteria = "Adult patients (≥18 years) with histologically confirmed diagnosis; excluded if prior cancer treatment within 6 months",
  outcome_definition = "Pathologically confirmed cancer status determined by independent review board",
  sample_size_justification = "Based on events per variable rule (EPV ≥10) and anticipated AUC of 0.80, required minimum 150 cases",
  missing_data_handling = "Multiple imputation using predictive mean matching (5 imputations)",
  background = "MicroRNAs are promising biomarkers but existing signatures lack validation",
  funding = "National Cancer Institute Grant R01-CA123456"
)

# Generate comprehensive report
report <- OmicSelector_tripod_report(
  model_result = result,
  study_info = study_info,
  output_format = "html",
  output_file = "tripod_compliance_report.html",
  include_plots = TRUE
)

# View checklist
print(report$checklist)
```

# PROBAST+AI Risk of Bias Assessment

## Run PROBAST Assessment

```{r}
# Provide assessment information
assessment_inputs <- list(
  participant_selection = "Consecutive enrollment of eligible patients",
  predictor_assessment = "RNA-seq performed using standardized protocol, blinded to outcome",
  outcome_assessment = "Independent pathological review blinded to biomarker results",
  sample_selection = "Hospital-based sample, representative of target population"
)

# Run assessment
assessment <- OmicSelector_probast(
  model_result = result,
  assessment_inputs = assessment_inputs
)

# View results
print(assessment)

# Check recommendations
assessment$recommendations
```

# Advanced: Custom Feature Selection

## Stability Selection

```{r}
# Use stability selection for more robust feature selection
result_stable <- OmicSelector_nested_cv(
  data = miR_Asakura,
  outcome = "Class",
  outer_folds = 5,
  inner_folds = 5,
  preprocessing_recipe = preprocessing_recipe,
  feature_selection_method = "stability_selection",
  max_features = 15,
  models = models
)

# Compare feature stability
result_stable$feature_stability$stable_features
```

## LASSO-based Selection

```{r}
# Use LASSO for feature selection
result_lasso <- OmicSelector_nested_cv(
  data = miR_Asakura,
  outcome = "Class",
  outer_folds = 5,
  inner_folds = 5,
  preprocessing_recipe = preprocessing_recipe,
  feature_selection_method = "lasso",
  models = models
)
```

# Phase 2: Advanced Feature Selection with Stability Metrics

Phase 2 introduces stability-aware feature selection methods that address the reproducibility crisis in biomarker discovery.

## Problem: Unstable Feature Selection

Traditional feature selection methods often produce different features on slightly different data:

```{r}
# Traditional approach (unstable)
# Run 1: Features A, B, C, D, E
# Run 2: Features A, F, G, H, E
# Run 3: Features A, B, I, J, K
# Only feature A is consistent!
```

## Solution: Stability Selection

OmicSelector 2.0 includes stability-aware methods that only select features consistently chosen across multiple iterations:

### Basic Stability Selection

```{r}
# Stability selection with subsampling
stable_result <- OmicSelector_stable_features(
  data = miR_Asakura,
  outcome = "Class",
  method = "stability_selection",
  n_iterations = 100,            # Run 100 times
  selection_threshold = 0.7,     # Keep features selected ≥70% of time
  subsample_rate = 0.8,          # Use 80% of data each iteration
  parallel = TRUE,
  seed = 123
)

# View results
print(stable_result)

# Top stable features
head(stable_result$selected_features, 20)

# Stability scores for each feature
head(stable_result$stability_scores, 20)
```

### Nogueira Stability Metrics

The package implements comprehensive stability metrics from Nogueira & Brown (2016):

```{r}
# View stability metrics
stable_result$nogueira_metrics

# Kuncheva Index (higher is better, >0.8 is excellent)
# Values range from -1 to 1
stable_result$nogueira_metrics$kuncheva_index

# Mean Jaccard Similarity (0 to 1)
stable_result$nogueira_metrics$mean_jaccard

# Variance in number of features selected
stable_result$nogueira_metrics$variance_n_features
```

**Interpretation:**

- **Kuncheva Index > 0.8**: Very stable selection
- **Kuncheva Index 0.5-0.8**: Moderately stable
- **Kuncheva Index < 0.5**: Unstable (problematic!)

### Visualize Stability

```{r}
# Plot stability scores
plot(stable_result)

# This creates a bar plot showing:
# - Top features by stability score
# - Selection threshold line
# - Color-coded by selection status
```

### Different Stability Methods

```{r}
# Boruta with stability assessment
boruta_stable <- OmicSelector_stable_features(
  data = miR_Asakura,
  outcome = "Class",
  method = "boruta_stable",
  n_iterations = 50,
  selection_threshold = 0.6
)

# LASSO with stability
lasso_stable <- OmicSelector_stable_features(
  data = miR_Asakura,
  outcome = "Class",
  method = "lasso_stable",
  n_iterations = 100,
  selection_threshold = 0.7
)

# RFE with stability
rfe_stable <- OmicSelector_stable_features(
  data = miR_Asakura,
  outcome = "Class",
  method = "rfe_stable",
  n_iterations = 50,
  selection_threshold = 0.6
)

# Compare stability across methods
comparison <- data.frame(
  Method = c("Stability Selection", "Boruta Stable", "LASSO Stable", "RFE Stable"),
  Kuncheva = c(
    stable_result$nogueira_metrics$kuncheva_index,
    boruta_stable$nogueira_metrics$kuncheva_index,
    lasso_stable$nogueira_metrics$kuncheva_index,
    rfe_stable$nogueira_metrics$kuncheva_index
  ),
  N_Features = c(
    length(stable_result$selected_features),
    length(boruta_stable$selected_features),
    length(lasso_stable$selected_features),
    length(rfe_stable$selected_features)
  )
)

print(comparison)
```

## Feature Clustering for Biomarker Replaceability

What happens when your selected biomarker can't be measured on a new platform? Feature clustering identifies alternative biomarkers.

### Correlation-Based Clustering

```{r}
# Cluster highly correlated features
cluster_result <- OmicSelector_cluster_features(
  data = miR_Asakura[, -which(names(miR_Asakura) == "Class")],  # Remove outcome
  method = "correlation",
  min_correlation = 0.7,      # Group features with r ≥ 0.7
  verbose = TRUE
)

# View results
print(cluster_result)

# See representative features from each cluster
cluster_result$representatives

# See all cluster assignments
head(cluster_result$cluster_assignments, 20)
```

### Find Alternative Biomarkers

```{r}
# Suppose you can't measure hsa-miR-21-5p on your new platform
# Find alternatives from the same cluster

# Get cluster ID for this feature
feature_of_interest <- "hsa-miR-21-5p"
cluster_id <- cluster_result$cluster_assignments[feature_of_interest]

# Get all features in the same cluster
same_cluster <- names(cluster_result$cluster_assignments)[
  cluster_result$cluster_assignments == cluster_id
]

# These are your alternative biomarkers
alternatives <- setdiff(same_cluster, feature_of_interest)
print(alternatives)

# Check their correlation
cor(miR_Asakura[, c(feature_of_interest, alternatives)])
```

### Different Clustering Methods

```{r}
# Hierarchical clustering
cluster_hier <- OmicSelector_cluster_features(
  data = miR_Asakura[, -which(names(miR_Asakura) == "Class")],
  method = "hierarchical",
  n_clusters = 50,  # Specify number of clusters
  distance_metric = "pearson"
)

# Expression pattern clustering
cluster_pattern <- OmicSelector_cluster_features(
  data = miR_Asakura[, -which(names(miR_Asakura) == "Class")],
  method = "expression_pattern",
  n_clusters = NULL  # Auto-detect optimal number
)

# K-medoids clustering
cluster_kmed <- OmicSelector_cluster_features(
  data = miR_Asakura[, -which(names(miR_Asakura) == "Class")],
  method = "kmedoids",
  n_clusters = 40
)
```

### Dimensionality Reduction with Clustering

```{r}
# Reduce from 2,000+ features to ~50 representatives
cluster_result <- OmicSelector_cluster_features(
  data = miR_Asakura[, -which(names(miR_Asakura) == "Class")],
  method = "correlation",
  min_correlation = 0.75
)

# Original features
n_original <- ncol(miR_Asakura) - 1

# Representative features
n_representatives <- length(cluster_result$representatives)

# Reduction percentage
reduction_pct <- (1 - n_representatives / n_original) * 100
cat("Reduced features by", round(reduction_pct, 1), "%\n")

# Use representatives in downstream analysis
representative_data <- miR_Asakura[, c(cluster_result$representatives, "Class")]

# Train model on reduced data
model_reduced <- randomForest::randomForest(
  Class ~ .,
  data = representative_data
)
```

### Visualize Clustering

```{r}
# Plot clustering results
plot(cluster_result)

# This creates multiple visualizations:
# 1. Heatmap of correlation matrix with clusters
# 2. Dendrogram showing hierarchical relationships
# 3. Network plot of highly correlated features
```

## Feature Importance with Confidence Intervals

Phase 2 includes model-agnostic feature importance with proper handling of correlated features.

### Permutation Importance

```{r}
# First, train a model
library(randomForest)
model <- randomForest(Class ~ ., data = miR_Asakura)

# Calculate permutation importance with confidence intervals
importance_result <- OmicSelector_importance(
  model = model,
  data = miR_Asakura,
  outcome = "Class",
  method = "permutation",
  n_repeats = 20,              # Repeat 20 times for CI
  normalize = TRUE,            # Normalize to [0,1]
  parallel = TRUE
)

# View results
print(importance_result)

# Top 20 important features with confidence intervals
head(importance_result$importance, 20)

# Plot importance
plot(importance_result)
```

### Conditional Importance (Handles Correlations)

When features are correlated, standard permutation importance can be misleading. Conditional importance fixes this:

```{r}
# Calculate conditional importance
importance_cond <- OmicSelector_importance(
  model = model,
  data = miR_Asakura,
  outcome = "Class",
  method = "conditional",
  n_repeats = 10,
  conditional_grid = 5,        # Conditioning grid size
  parallel = TRUE
)

# Compare permutation vs conditional
importance_both <- OmicSelector_importance(
  model = model,
  data = miR_Asakura,
  outcome = "Class",
  method = "both",             # Calculate both methods
  n_repeats = 10
)

# See where they differ (indicates correlation effects)
comparison <- merge(
  importance_result$importance[, c("feature", "importance")],
  importance_cond$importance[, c("feature", "importance")],
  by = "feature",
  suffixes = c("_permutation", "_conditional")
)

# Features where conditional importance is much lower suggests
# they're important mainly due to correlation with other features
comparison$difference <- comparison$importance_permutation -
                         comparison$importance_conditional

# Biggest differences
head(comparison[order(-abs(comparison$difference)), ], 10)
```

### Integration: Stability + Clustering + Importance

Combine all Phase 2 methods for robust biomarker discovery:

```{r}
# Step 1: Stable feature selection
stable_features <- OmicSelector_stable_features(
  data = miR_Asakura,
  outcome = "Class",
  method = "stability_selection",
  n_iterations = 100,
  selection_threshold = 0.7
)

# Step 2: Cluster the stable features
stable_data <- miR_Asakura[, c(stable_features$selected_features, "Class")]
cluster_stable <- OmicSelector_cluster_features(
  data = stable_data[, -ncol(stable_data)],
  method = "correlation",
  min_correlation = 0.7
)

# Step 3: Train model on cluster representatives
representative_data <- stable_data[, c(cluster_stable$representatives, "Class")]
final_model <- randomForest(Class ~ ., data = representative_data)

# Step 4: Calculate importance
final_importance <- OmicSelector_importance(
  model = final_model,
  data = representative_data,
  outcome = "Class",
  method = "conditional",
  n_repeats = 20
)

# Final signature: Top features that are:
# 1. Stable (selected consistently)
# 2. Non-redundant (cluster representatives)
# 3. Important (high conditional importance)
final_signature <- head(final_importance$importance$feature, 10)

cat("Final robust signature (", length(final_signature), " features):\n")
print(final_signature)
```

### Summary Statistics

```{r}
# Create comprehensive summary
summary_stats <- list(
  original_features = ncol(miR_Asakura) - 1,
  stable_features = length(stable_features$selected_features),
  stability_kuncheva = stable_features$nogueira_metrics$kuncheva_index,
  clusters = length(unique(cluster_stable$cluster_assignments)),
  representatives = length(cluster_stable$representatives),
  final_signature = length(final_signature),
  reduction_pct = (1 - length(final_signature) / (ncol(miR_Asakura) - 1)) * 100
)

print(summary_stats)
```

# Comparison with Traditional Approach

## ❌ Old Way (Potential Leakage)

```{r}
# DON'T DO THIS - Shows what NOT to do
library(caret)

# 1. Normalize entire dataset (LEAKAGE!)
normalized_data <- preProcess(miR_Asakura, method = c("center", "scale"))

# 2. Select features on entire dataset (LEAKAGE!)
feature_selected <- Boruta(Class ~ ., data = normalized_data)

# 3. Split into train/test
train_indices <- createDataPartition(normalized_data$Class, p = 0.8)

# Performance will be optimistically biased!
```

## ✅ New Way (No Leakage)

```{r}
# OmicSelector 2.0 ensures all steps happen inside folds
result <- OmicSelector_nested_cv(
  data = miR_Asakura,  # Raw data
  outcome = "Class",
  preprocessing_recipe = preprocessing_recipe,  # Applied inside folds
  feature_selection_method = "boruta",          # Applied inside folds
  outer_folds = 5,
  inner_folds = 5
)

# Performance is unbiased!
```

# Single Model Training (Simpler Workflow)

If you don't need nested CV and want to train a single model:

```{r}
# Simple model training with automatic framework detection
simple_result <- OmicSelector_fit(
  data = miR_Asakura,
  outcome = "Class",
  method = "tidymodels",  # or "caret" for backward compatibility
  algorithm = "ranger",
  preprocessing = list(
    normalize = TRUE,
    remove_zero_variance = TRUE,
    impute = TRUE
  ),
  resampling = list(
    method = "cv",
    outer_folds = 10
  ),
  parallel = TRUE
)

# View results
print(simple_result)
summary(simple_result)
```

# Exporting Results

## Save Model Object

```{r}
# Save complete results
saveRDS(result, "nested_cv_results.rds")

# Load later
loaded_result <- readRDS("nested_cv_results.rds")
```

## Export Predictions

```{r}
# Extract predictions
predictions <- result$final_predictions

# Save as CSV
write.csv(predictions, "predictions.csv", row.names = FALSE)
```

## Export Selected Features

```{r}
# Get features from all folds
all_features <- result$selected_features

# Most stable features
stable_features <- result$feature_stability$stable_features

# Save
write.csv(
  data.frame(Feature = stable_features),
  "stable_features.csv",
  row.names = FALSE
)
```

# Best Practices

1. **Always use nested CV** for unbiased performance estimates
2. **Never preprocess before splitting** - use recipes inside folds
3. **Check feature stability** - prefer features selected consistently
4. **Generate TRIPOD+AI reports** for clinical models
5. **Assess risk of bias** using PROBAST+AI
6. **Set random seeds** for reproducibility
7. **Use parallel processing** for faster computation
8. **Validate externally** when possible

# Troubleshooting

## Common Issues

### Issue: Out of Memory

```{r}
# Solution: Reduce parallel processing or grid size
result <- OmicSelector_nested_cv(
  ...,
  parallel = TRUE,
  cores = 2,  # Use fewer cores
  tune_grid = 5  # Try fewer hyperparameter combinations
)
```

### Issue: Taking Too Long

```{r}
# Solution: Use fewer folds or smaller grid
result <- OmicSelector_nested_cv(
  ...,
  outer_folds = 3,  # Reduce from 5
  inner_folds = 3,  # Reduce from 5
  tune_grid = 5     # Reduce from 10
)
```

### Issue: Feature Selection Finds Nothing

```{r}
# Solution: Relax selection criteria
result <- OmicSelector_nested_cv(
  ...,
  feature_selection_method = "lasso",  # Try different method
  max_features = 50  # Allow more features
)
```

# Next Steps

- Read the full documentation: `?OmicSelector_nested_cv`
- Explore Phase 2 features: Multi-omics integration
- Check out survival analysis capabilities (Phase 5)
- Try AutoML features (Phase 7)

# Session Info

```{r}
sessionInfo()
```
