---
title: "Single-sample deployment with OmicSelector"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-sample deployment with OmicSelector}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(OmicSelector)
```

## Motivation

> **Deployment problem.** Classical batch-aware corrections such as ComBat,
> quantile normalization, or RUV are designed around a batch or reference set.
> That is a mismatch for a clinical-style handoff where one new specimen arrives
> alone and must be scored without co-resident test samples.
>
> **Honest caveat.** The value of the API below is *deployability*: it scores
> one incoming specimen from its own panel values plus frozen fit-time
> parameters, with no co-resident test batch. It is not a claim that
> `ws-balance-ilr`, or any single-sample method, out-discriminates a
> batch-corrected pipeline — choose and validate the scoring method on your own
> data.

## Make a tiny synthetic compositional panel

This vignette uses a self-contained synthetic count-like matrix. The feature
names come from the default within-sample balance dictionary so the example
exercises the same panel shape as the deployment tests, but no data are
downloaded.

```{r synthetic-data}
set.seed(101)

sbp <- ws_default_sbp()
sentinels <- c("ALL_NON_RBC", "ALL_NON_PLT",
               "TOP_K_BY_ABUNDANCE", "TAIL_BY_ABUNDANCE")
features <- unique(unlist(lapply(sbp, function(balance) {
  c(balance$numerator, balance$denominator)
}), use.names = FALSE))
features <- features[!features %in% sentinels]
features <- features[seq_len(30L)]

n <- 40L
y <- rep(c(0L, 1L), each = n / 2L)
X <- matrix(
  stats::rgamma(n * length(features), shape = 20, rate = 2),
  nrow = n,
  dimnames = list(paste0("S", seq_len(n)), features)
)
X[y == 1L, features[1:5]] <- X[y == 1L, features[1:5]] + 80
X[y == 0L, features[6:10]] <- X[y == 0L, features[6:10]] + 80

holdout <- nrow(X)
X_train <- X[-holdout, , drop = FALSE]
y_train <- y[-holdout]
x_new <- X[holdout, , drop = FALSE]

dim(X_train)
dim(x_new)
```

## Fit once, then freeze the scorer

`deploy_singlesample()` fits one rostered method and stores the frozen model
state needed for future singleton scoring. The default `ws-balance-ilr` method
is a compositional deployment default, not a certified benchmark winner.

```{r fit-freeze}
dep <- deploy_singlesample(
  X_train,
  y_train,
  method = "ws-balance-ilr"
)

print(dep)
```

## Score one incoming specimen

`score_specimen()` needs only the held-out specimen row and the frozen deployment
object. No test batch, co-resident reference cohort, or test-time batch
correction is supplied.

```{r score-one}
score <- score_specimen(dep, x_new)
score
```

## Check the row-equivariance guarantee

The deployability gate checks the property needed for singleton use: a
specimen's score is the same when computed alone as when computed as a row in a
batch, with the frozen model unchanged.

```{r deployability-gate}
is_singlesample_deployable(dep, X_probe = X_train)
```

## Inspect the method roster, not a ranking

The deployment wrapper is intentionally fail-closed. Roughly 30 rostered
single-sample methods are deployable through this fit-freeze-score pattern, and
methods needing cohort context, unavailable backends, or non-scalar outputs
reject cleanly rather than pretending to be singleton-safe. Use
`singlesample_method_roster()` to inspect the current roster and do not treat
the roster order as a performance ranking.

```{r roster}
roster <- singlesample_method_roster()
head(roster[, c("method_id", "tier", "dep_route")], 8)
```

The deployability contract is what the roster guarantees, not a ranking: a
deployable method can score one incoming specimen without a new batch or a
test-time batch-correction step. It does not certify that any method provides
superior discrimination, so benchmark the method choice on your own cohorts.
