---
title: "Biomarker Discovery with TCGA Data"
subtitle: "A Complete Workflow Using OmicSelector 2.0"
author: "OmicSelector Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Biomarker Discovery with TCGA Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This tutorial demonstrates a complete biomarker discovery workflow using **OmicSelector 2.0** with real data from The Cancer Genome Atlas (TCGA). We'll use miRNA expression data to discover diagnostic biomarkers for pancreatic cancer.

## Learning Objectives

By the end of this tutorial, you will be able to:

1. Prepare TCGA data for biomarker discovery
2. Configure and run feature selection with zero data leakage
3. Compare multiple filter-model combinations using nested CV
4. Analyze feature stability across cross-validation folds
5. Select an optimal biomarker signature balancing performance and stability
6. Train and save a final model for deployment
7. Apply batch correction using FrozenComBat
8. Assess and improve model calibration
9. Generate a TRIPOD+AI compliant report

## Prerequisites

```{r prerequisites}
# Install OmicSelector if needed
if (!require("OmicSelector")) {
  remotes::install_github("kstawiski/OmicSelector")
}

# Load required packages
library(OmicSelector)
library(mlr3)
library(mlr3learners)
library(data.table)
library(ggplot2)

# For TCGA data download (optional)
# BiocManager::install("TCGAbiolinks")
```

---

# Part 1: Data Preparation

## 1.1 Loading TCGA Data

OmicSelector includes pre-processed TCGA miRNA expression data for demonstration.

```{r load-data}
# Load packaged TCGA data
data("orginal_TCGA_data", package = "OmicSelector")

# Examine the data structure
cat("Dataset dimensions:", nrow(orginal_TCGA_data), "samples x",
    ncol(orginal_TCGA_data), "columns\n")

# View available cancer types
table(orginal_TCGA_data$primary_site)
```

## 1.2 Selecting Cases and Controls

For a valid biomarker study, we need properly matched cases and controls.
Using tumor vs. normal tissue from the **same organ** is critical.

```{r select-samples}
# Select pancreatic cancer cases (primary tumors only)
pancreatic_cancer <- orginal_TCGA_data[
  orginal_TCGA_data$primary_site == "Pancreas" &
  orginal_TCGA_data$sample_type == "PrimaryTumor",
]

# Select pancreatic normal controls
pancreatic_normal <- orginal_TCGA_data[
  orginal_TCGA_data$primary_site == "Pancreas" &
  orginal_TCGA_data$sample_type == "SolidTissueNormal",
]

cat("Pancreatic cancer samples:", nrow(pancreatic_cancer), "\n")
cat("Pancreatic normal samples:", nrow(pancreatic_normal), "\n")
```

**Important**: Never mix normal tissues from different organs as controls - this creates confounding that leads to inflated performance estimates.

## 1.3 Handling Limited Normal Samples

TCGA often has few matched normal samples. We can carefully add normals from related tissues.

**⚠️ IMPORTANT WARNING**: Adding normal tissues from different organs introduces
**biological confounding**. Expression differences between organs may dominate
tumor vs. normal differences, potentially leading to:

- **Inflated performance estimates** - You may be classifying "pancreas vs. colon"
  rather than "tumor vs. normal"
- **Biologically meaningless biomarkers** - Features distinguishing tissue types,
  not cancer status
- **Failed external validation** - Models trained this way often fail when tested
  on matched tumor-normal pairs

**Recommendations:**
1. **Preferred**: Use only matched tissue normals, even if sample size is small
2. **Alternative**: Use an independent normal cohort from the same tissue
3. **Last resort**: Add related tissue normals, but **clearly report this limitation**
   and interpret results with extreme caution

```{r handle-limited-normals}
# If we have fewer than 30 pancreatic normals, add GI tract normals
# WARNING: This is a last resort that introduces biological confounding!
if (nrow(pancreatic_normal) < 30) {
  warning("Adding GI tract normals - this introduces biological confounding!")
  message("Adding GI tract normals to increase control sample size")
  message("Results should be validated on matched tumor-normal pairs.")

  # Add normal tissues from related GI organs
  gi_organs <- c("Stomach", "Colon", "Liver and intrahepatic bile ducts")
  gi_normals <- orginal_TCGA_data[
    orginal_TCGA_data$primary_site %in% gi_organs &
    orginal_TCGA_data$sample_type == "SolidTissueNormal",
  ]

  # Sample to avoid overwhelming imbalance
  set.seed(42)
  n_to_add <- min(50, nrow(gi_normals))
  gi_sample <- gi_normals[sample(nrow(gi_normals), n_to_add), ]

  pancreatic_normal <- rbind(pancreatic_normal, gi_sample)
  cat("Total controls after augmentation:", nrow(pancreatic_normal), "\n")
  cat("WARNING: These results may be confounded by tissue-type differences.\n")
}
```

## 1.4 Creating the Analysis Dataset

```{r create-dataset}
# Add class labels
pancreatic_cancer$Class <- "Case"
pancreatic_normal$Class <- "Control"

# Combine into final dataset
dataset <- rbind(pancreatic_cancer, pancreatic_normal)

# Extract miRNA features (columns starting with "hsa")
mirna_cols <- grep("^hsa", names(dataset), value = TRUE)
cat("Number of miRNA features:", length(mirna_cols), "\n")

# Create analysis data.frame
# Handle both data.table and data.frame inputs
if (inherits(dataset, "data.table")) {
  mirna_matrix <- as.data.frame(dataset[, mirna_cols, with = FALSE])
} else {
  mirna_matrix <- dataset[, mirna_cols]
}

analysis_data <- data.frame(
  mirna_matrix,
  outcome = factor(dataset$Class, levels = c("Control", "Case"))
)

# Check class distribution
table(analysis_data$outcome)
```

## 1.5 Data Quality Checks

Before analysis, always perform quality checks:

```{r quality-checks}
# Check for missing values
na_count <- sum(is.na(analysis_data))
cat("Total missing values:", na_count, "\n")

# Check feature variance
feature_vars <- apply(analysis_data[, mirna_cols], 2, var, na.rm = TRUE)
low_var_features <- sum(feature_vars < 0.01)
cat("Features with very low variance:", low_var_features, "\n")

# Remove zero-variance features
valid_features <- names(feature_vars[feature_vars > 0])
analysis_data <- analysis_data[, c(valid_features, "outcome")]

cat("Features after QC:", length(valid_features), "\n")
```

---

# Part 2: Building the Analysis Pipeline

## 2.1 Creating an OmicPipeline

```{r create-pipeline}
# Create pipeline
pipeline <- OmicPipeline$new(
  data = analysis_data,
  target = "outcome",
  positive = "Case"    # Define which class is "positive" for AUC
)

# View pipeline summary
print(pipeline)
```

## 2.2 Understanding GraphLearners

OmicSelector uses `mlr3pipelines` GraphLearners to ensure all preprocessing
happens **inside** cross-validation folds. The pipeline order is:

```
Impute → Scale → [Batch Correct] → [Oversample] → Filter → Model
                   optional         optional
```

```{r create-learner}
# Create a basic GraphLearner
learner_basic <- pipeline$create_graph_learner(
  filter = "anova",           # ANOVA F-test for feature selection
  model = "ranger",           # Random Forest
  n_features = 30,            # Select top 30 features
  impute_method = "median",   # Impute missing values with median
  scale = TRUE                # Standardize features
)

# View the learner structure
print(learner_basic)
```

## 2.3 Available Options

### Feature Selection Methods

| Filter | Description | Strengths |
|--------|-------------|-----------|
| `"anova"` | ANOVA F-test | Good default, tests mean differences |
| `"variance"` | Variance filter | Removes uninformative features |
| `"correlation"` | Target correlation | Fast univariate filter |
| `"mrmr"` | Min Redundancy Max Relevance | Reduces redundant features |
| `"information_gain"` | Entropy-based | Good for mixed data types |

### Classification Models

| Model | Description | When to Use |
|-------|-------------|-------------|
| `"ranger"` | Random Forest | Default choice, handles interactions |
| `"glmnet"` | Elastic Net | When interpretability is important |
| `"svm"` | Support Vector Machine | High-dimensional data |
| `"log_reg"` | Logistic Regression | Baseline comparison |

---

# Part 3: Comparing Multiple Configurations

## 3.1 Defining a Comparison Grid

We'll compare different filter-model combinations:

```{r define-grid}
# Define configurations to compare
configs <- list(
  list(filter = "anova", model = "ranger", n_features = 20, id = "anova_rf_20"),
  list(filter = "anova", model = "ranger", n_features = 30, id = "anova_rf_30"),
  list(filter = "anova", model = "glmnet", n_features = 30, id = "anova_glmnet_30"),
  list(filter = "mrmr", model = "ranger", n_features = 20, id = "mrmr_rf_20"),
  list(filter = "variance", model = "svm", n_features = 30, id = "var_svm_30")
)

cat("Testing", length(configs), "configurations\n")
```

## 3.2 Setting Up the Benchmark

```{r setup-benchmark}
# Create benchmark service with nested cross-validation
benchmark <- BenchmarkService$new(
  task = pipeline,
  outer_folds = 5,    # 5-fold outer CV for evaluation
  inner_folds = 3,    # 3-fold inner CV for feature selection
  seed = 42           # For reproducibility
)

# Add all configurations
for (cfg in configs) {
  learner <- pipeline$create_graph_learner(
    filter = cfg$filter,
    model = cfg$model,
    n_features = cfg$n_features
  )
  benchmark$add_learner(learner, id = cfg$id)
}

print(benchmark)
```

## 3.3 Running the Benchmark

```{r run-benchmark}
# Run nested cross-validation (this may take several minutes)
cat("Starting nested cross-validation...\n")
cat("This may take 5-10 minutes depending on data size.\n\n")

result <- benchmark$run()

# View performance summary
print(result)
```

## 3.4 Analyzing Results

```{r analyze-results}
# Get performance table
performance <- result$performance
print(performance)

# Compute standard errors from resampling (if available)
# The benchmark stores per-fold scores that we can use
bmr <- result$benchmark_result
fold_scores <- bmr$score(msr("classif.auc"))

# Aggregate mean and standard error per learner
library(data.table)
scores_dt <- as.data.table(fold_scores)
perf_summary <- scores_dt[, .(
  mean_auc = mean(classif.auc, na.rm = TRUE),
  se_auc = sd(classif.auc, na.rm = TRUE) / sqrt(.N)
), by = learner_id]

# Visualize performance with proper error bars
ggplot(perf_summary, aes(x = reorder(learner_id, mean_auc),
                          y = mean_auc)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_auc - se_auc, ymax = mean_auc + se_auc),
                width = 0.2) +
  coord_flip() +
  labs(
    title = "Performance Comparison: AUC",
    subtitle = "Error bars show standard error across CV folds",
    x = "Configuration",
    y = "AUC (Cross-Validated)"
  ) +
  theme_minimal()
```

---

# Part 4: Feature Stability Analysis

## 4.1 Why Stability Matters

High performance with unstable features often indicates overfitting.
The **Nogueira Stability Index** quantifies feature selection consistency:

- **SI ≥ 0.9**: Excellent - Very stable selections
- **SI 0.7-0.9**: Good - Reasonably stable
- **SI 0.5-0.7**: Moderate - Some instability
- **SI < 0.5**: Poor - Likely overfitting

## 4.2 Computing Stability

```{r compute-stability}
# Compute stability for the benchmark result
stability <- compute_stability_from_resample(result$benchmark_result)

print(stability)

# View interpretation
cat("\n", stability$interpretation, "\n")
```

## 4.3 Examining Consensus Features

Features selected in ALL folds (consensus) are most reliable:

```{r consensus-features}
# Get consensus features
consensus <- stability$consensus_features
cat("Consensus features (selected in all folds):", length(consensus), "\n")
print(head(consensus, 20))

# View selection frequency
freq <- sort(stability$selection_frequency, decreasing = TRUE)
head(freq, 20)
```

## 4.4 Visualizing Feature Stability

```{r plot-stability}
# Create selection frequency plot
freq_df <- data.frame(
  feature = names(freq)[1:20],
  frequency = freq[1:20]
)
freq_df$feature <- factor(freq_df$feature, levels = rev(freq_df$feature))

ggplot(freq_df, aes(x = frequency, y = feature)) +
  geom_col(aes(fill = frequency >= 1.0), alpha = 0.8) +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "darkgreen"),
    labels = c("Variable", "Consensus"),
    name = "Selection"
  ) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "red") +
  labs(
    title = "Top 20 Features by Selection Frequency",
    x = "Selection Frequency (1.0 = all folds)",
    y = NULL
  ) +
  theme_minimal()
```

---

# Part 5: Optimal Signature Selection

## 5.1 Multi-Objective Selection

OmicSelector selects signatures balancing:
- **Performance** (AUC)
- **Stability** (Nogueira Index)
- **Parsimony** (fewer features)

```{r select-signature}
# Select best signature using weighted mode
best <- select_best_signature(
  result,
  mode = "weighted",
  stability_weight = 0.3,    # 30% weight on stability
  parsimony_weight = 0.1     # 10% weight on fewer features
)

print(best)
```

## 5.2 Selection Modes

| Mode | Description | Use When |
|------|-------------|----------|
| `"best_auc"` | Highest AUC only | Maximum predictive power needed |
| `"best_stability"` | Highest stability only | Reproducibility is critical |
| `"weighted"` | Weighted combination | Balanced selection (default) |
| `"pareto"` | Pareto-optimal with utopia | Formal multi-objective |

## 5.3 Final Signature

```{r final-signature}
# Extract the final signature
final_features <- best$features
cat("Final signature contains", length(final_features), "features:\n")
print(final_features)

# Summary metrics
cat("\nSignature Metrics:\n")
cat("  AUC:", round(best$auc, 3), "\n")
cat("  Stability:", round(best$stability, 3), "\n")
cat("  N Features:", length(final_features), "\n")
```

---

# Part 6: Training the Final Model

After selecting an optimal signature configuration, we train a final model on the
**full dataset** for deployment or external validation.

## 6.1 Selecting the Best Learner

```{r select-best-learner}
# From benchmarking, identify the best configuration
# Let's say "anova_rf_30" performed best (example)
best_config <- "anova_rf_30"

# Create the learner with the optimal configuration
final_learner <- pipeline$create_graph_learner(
  filter = "anova",
  model = "ranger",
  n_features = 30,
  scale = TRUE
)

# Set the learner ID
final_learner$id <- "final_model"
```

## 6.2 Training on Full Data

```{r train-final-model}
# Get the task from the pipeline
task <- pipeline$get_task()

# Train the final model on ALL data
# (Feature selection happens inside the pipeline, so no leakage)
final_learner$train(task)

cat("Final model trained successfully.\n")
cat("Model type:", final_learner$model$model, "\n")

# Extract selected features from the trained pipeline
# The filter step stores which features were selected
selected_features <- final_learner$state$model$filter$features
cat("Features selected:", length(selected_features), "\n")
print(head(selected_features, 10))
```

## 6.3 Generating Predictions

```{r final-predictions}
# Generate predictions on the training data (for calibration)
predictions <- final_learner$predict(task)

# Extract probabilities for the positive class
prob_positive <- predictions$prob[, "Case"]

# Extract true labels
true_labels <- as.integer(task$truth() == "Case")

# Summary statistics
cat("Prediction summary:\n")
cat("  Mean probability:", round(mean(prob_positive), 3), "\n")
cat("  Min:", round(min(prob_positive), 3), "\n")
cat("  Max:", round(max(prob_positive), 3), "\n")
cat("  AUC (resubstitution):",
    round(predictions$score(msr("classif.auc")), 3), "\n")
```

**Note**: The resubstitution AUC is optimistically biased. The nested CV results
from Part 3 provide unbiased performance estimates.

## 6.4 Saving the Model

```{r save-model}
# Save the trained model for later use
saveRDS(final_learner, "final_pancreatic_model.rds")
cat("Model saved to 'final_pancreatic_model.rds'\n")

# To load later:
# loaded_model <- readRDS("final_pancreatic_model.rds")
# new_predictions <- loaded_model$predict_newdata(new_data)
```

---

# Part 7: Batch Correction with FrozenComBat

## 7.1 Why FrozenComBat?

When applying a model to external data with batch effects, we must:
1. Fit batch parameters on **training data only**
2. Apply the same transformation to new data

Standard ComBat re-fits on all data, causing **leakage**.
**FrozenComBat** solves this by "freezing" parameters from training.

## 7.2 Creating a Batch Column

```{r create-batch}
# Simulate batch information (in practice, extract from sample metadata)
set.seed(42)
analysis_data$batch <- sample(
  c("Batch_A", "Batch_B", "Batch_C"),
  nrow(analysis_data),
  replace = TRUE
)

# View batch distribution
table(analysis_data$batch, analysis_data$outcome)
```

## 7.3 Including Batch Correction in Pipeline

```{r batch-correction}
# Create pipeline with batch information
pipeline_bc <- OmicPipeline$new(
  data = analysis_data,
  target = "outcome",
  positive = "Case",
  batch = "batch"     # Specify batch column
)

# Create learner with batch correction enabled
learner_bc <- pipeline_bc$create_graph_learner(
  filter = "anova",
  model = "ranger",
  n_features = 30,
  batch_correct = TRUE   # Enable FrozenComBat
)

print(learner_bc)
```

## 7.4 Verifying No Batch Leakage

```{r check-batch-leakage}
# After running benchmark, check for batch effect reduction
# If reduction > 95%, this may indicate leakage
check_result <- check_batch_correction_leakage(
  data = analysis_data[, mirna_cols[1:50]],  # Sample of features
  batch = analysis_data$batch
)

print(check_result)
```

---

# Part 8: Model Calibration

## 8.1 Why Calibrate?

Predicted probabilities may not reflect true event rates.
Calibration ensures `P(outcome|prediction=0.7) ≈ 70%`.

**When to calibrate:**
- When probability estimates (not just classifications) matter
- For risk stratification or clinical decision support
- When combining predictions with other probability-based models

## 8.2 Computing Calibration Metrics

Using the predictions from Part 6 (final model training):

```{r calibration-metrics}
# We already have prob_positive and true_labels from Part 6
# prob_positive <- predictions$prob[, "Case"]
# true_labels <- as.integer(task$truth() == "Case")

# Compute Expected Calibration Error (ECE)
# ECE measures how well predicted probabilities match observed frequencies
ece <- compute_ece(prob_positive, true_labels)
cat("Expected Calibration Error (ECE):", round(ece, 4), "\n")

# Interpretation:
# ECE < 0.05: Well-calibrated
# ECE 0.05-0.10: Acceptable
# ECE > 0.10: Consider recalibration

# Decompose Brier Score into reliability, resolution, uncertainty
brier <- decompose_brier(prob_positive, true_labels)
cat("\nBrier Score Decomposition:\n")
cat("  Total Brier Score:", round(brier$brier, 4), "\n")
cat("  Reliability (lower is better):", round(brier$reliability, 4), "\n")
cat("  Resolution (higher is better):", round(brier$resolution, 4), "\n")
cat("  Uncertainty (baseline):", round(brier$uncertainty, 4), "\n")
```

## 8.3 Calibration Plot

```{r calibration-plot}
# Create calibration curve data
n_bins <- 10
breaks <- seq(0, 1, length.out = n_bins + 1)
bins <- cut(prob_positive, breaks = breaks, include.lowest = TRUE)

cal_data <- data.frame(
  bin = levels(bins),
  predicted = tapply(prob_positive, bins, mean),
  observed = tapply(true_labels, bins, mean),
  n = as.numeric(table(bins))
)
cal_data <- cal_data[!is.na(cal_data$predicted), ]

# Plot calibration curve
ggplot(cal_data, aes(x = predicted, y = observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(size = n), color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "darkblue") +
  scale_size_continuous(name = "N samples", range = c(2, 8)) +
  labs(
    title = "Calibration Plot",
    subtitle = paste("ECE =", round(ece, 3)),
    x = "Mean Predicted Probability",
    y = "Observed Frequency"
  ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal()
```

## 8.4 Applying Calibration

If calibration is poor (ECE > 0.10), apply recalibration:

```{r apply-calibration}
# Option 1: Platt Scaling (logistic recalibration)
# Best for sigmoid-shaped miscalibration
calibrator_platt <- fit_platt_scaling(prob_positive, true_labels)

# Option 2: Isotonic Regression
# More flexible, best for non-monotonic miscalibration
calibrator_isotonic <- fit_isotonic_calibration(prob_positive, true_labels)

# Apply calibration to the same data (for demonstration)
calibrated_platt <- calibrator_platt$calibrate(prob_positive)
calibrated_isotonic <- calibrator_isotonic$calibrate(prob_positive)

# Compare calibration errors
ece_original <- compute_ece(prob_positive, true_labels)
ece_platt <- compute_ece(calibrated_platt, true_labels)
ece_isotonic <- compute_ece(calibrated_isotonic, true_labels)

cat("Calibration Comparison:\n")
cat("  Original ECE:", round(ece_original, 4), "\n")
cat("  After Platt Scaling:", round(ece_platt, 4), "\n")
cat("  After Isotonic Regression:", round(ece_isotonic, 4), "\n")

# Save the calibrator for use with new predictions
saveRDS(calibrator_platt, "calibrator_platt.rds")
cat("\nCalibrator saved for future use.\n")
```

**Important**: For rigorous calibration:
- Use a held-out calibration set, not the training data
- Or perform calibration within cross-validation folds

---

# Part 9: Generating Reports

## 9.1 TRIPOD+AI Compliant Report

OmicSelector generates reports following the TRIPOD+AI checklist:

```{r generate-report}
# Generate comprehensive report
generate_tripod_report(
  results = result,
  output_file = "tcga_pancreatic_report.html",
  title = "Pancreatic Cancer miRNA Biomarker Discovery",
  author = "OmicSelector Analysis"
)
```

## 9.2 Report Contents

The generated report includes:

1. **Executive Summary** - Key metrics at a glance
2. **Methods** - Data source, validation strategy, model details
3. **Discrimination** - AUC, sensitivity, specificity
4. **Calibration** - Brier score, calibration plot
5. **Stability** - Nogueira index, consensus features
6. **TRIPOD Checklist** - Compliance status

---

# Summary

In this tutorial, we demonstrated:

1. **Data Preparation**: Loading TCGA data, selecting proper controls
2. **Pipeline Creation**: Using OmicPipeline with GraphLearners
3. **Benchmarking**: Comparing multiple configurations with nested CV
4. **Stability Analysis**: Computing and interpreting Nogueira Index
5. **Signature Selection**: Multi-objective selection of optimal features
6. **Final Model Training**: Training on full data for deployment
7. **Batch Correction**: FrozenComBat for leakage-free batch correction
8. **Calibration**: Ensuring reliable probability predictions
9. **Reporting**: TRIPOD+AI compliant documentation

## Key Takeaways

- Always use **nested cross-validation** for unbiased performance estimates
- Check **feature stability** - high AUC with low stability = overfitting
- Use **consensus features** (selected in all folds) for robust signatures
- Apply **FrozenComBat** when external validation involves batch effects
- Generate **TRIPOD+AI reports** for publication-ready documentation

## Next Steps

- Try different feature selection methods
- Explore multi-omics integration
- Apply the signature to external validation cohorts
- Use SHAP values to interpret feature importance

---

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
