---
title: "OmicSelector 2.0 - TRIPOD+AI Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
params:
  report_data: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)

# Access report data
rd <- params$report_data
```

# Executive Summary

This report documents a biomarker discovery analysis performed with **OmicSelector 2.0**, adhering to TRIPOD+AI (Transparent Reporting of a multivariable prediction model for Individual Prognosis Or Diagnosis - Artificial Intelligence) guidelines.

**Analysis Timestamp:** `r format(rd$meta$timestamp, '%Y-%m-%d %H:%M:%S')`

**Key Results:**

| Metric | Value |
|--------|-------|
| Samples | `r rd$data$n_samples %||% "N/A"` |
| Features | `r rd$data$n_features %||% "N/A"` |
| AUC | `r sprintf("%.3f", rd$discrimination$auc %||% NA)` |
| Brier Score | `r sprintf("%.3f", rd$discrimination$brier_score %||% NA)` |
| Stability (Nogueira) | `r sprintf("%.3f", rd$stability$nogueira_index %||% NA)` |

---

# TRIPOD Checklist

## Title and Abstract (Items 1-2)

- **Study Type:** Prediction model development with internal validation
- **Target Outcome:** `r rd$data$target_name %||% "Not specified"`
- **Modeling Approach:** Machine learning pipeline with nested cross-validation

## Methods

### Source of Data (Item 4)

```{r data-summary}
data_info <- data.frame(
  Characteristic = c("Total Samples", "Total Features", "Modalities", "Target Variable"),
  Value = c(
    rd$data$n_samples %||% "N/A",
    rd$data$n_features %||% "N/A",
    rd$data$n_modalities %||% "1",
    rd$data$target_name %||% "N/A"
  )
)
knitr::kable(data_info, caption = "Data Characteristics")
```

### Outcome (Item 6)

**Class Distribution:**

```{r class-dist, eval=!is.null(rd$data$class_distribution)}
if (!is.null(rd$data$class_distribution)) {
  class_df <- as.data.frame(rd$data$class_distribution)
  names(class_df) <- c("Class", "Count")
  knitr::kable(class_df, caption = "Outcome Class Distribution")
}
```

### Predictors (Item 7)

- **Feature Selection Method:** `r rd$model$filter_method %||% "Not specified"`
- **Number of Features Selected:** `r rd$model$n_features_selected %||% "Variable"`
- **Model Type:** `r rd$model$learner_type %||% "Not specified"`

### Sample Size (Item 8)

- **Total samples:** `r rd$data$n_samples %||% "N/A"`
- **Events per variable (EPV):** Computed based on minority class and selected features

### Missing Data (Item 9)

```{r missing-data}
if (!is.null(rd$data$missing_data)) {
  cat(sprintf("Missing values: %d (%.1f%%)",
              rd$data$missing_data$count,
              rd$data$missing_data$percent))
} else {
  cat("Missing data handling: Median imputation within CV folds")
}
```

### Model Development and Validation (Item 10)

**Validation Strategy:**

- Type: `r rd$validation$type %||% "Internal nested cross-validation"`
- Outer Folds: `r rd$validation$outer_folds %||% "N/A"`
- Inner Folds: `r rd$validation$inner_folds %||% "N/A"`
- Stratified: `r ifelse(isTRUE(rd$validation$stratified), "Yes", "No")`

**Zero-Leakage Guarantee:** All preprocessing, feature selection, and model training occur within CV folds via mlr3 GraphLearners.

---

# Results

## Discrimination (Item 16)

```{r discrimination-table}
disc_metrics <- data.frame(
  Metric = c("AUC (ROC)", "Brier Score", "Sensitivity", "Specificity", "PPV", "NPV"),
  Value = c(
    sprintf("%.3f", rd$discrimination$auc %||% NA),
    sprintf("%.3f", rd$discrimination$brier_score %||% NA),
    sprintf("%.3f", rd$discrimination$sensitivity %||% NA),
    sprintf("%.3f", rd$discrimination$specificity %||% NA),
    sprintf("%.3f", rd$discrimination$ppv %||% NA),
    sprintf("%.3f", rd$discrimination$npv %||% NA)
  )
)
disc_metrics <- disc_metrics[!is.na(disc_metrics$Value) & disc_metrics$Value != "NA", ]
knitr::kable(disc_metrics, caption = "Discrimination Metrics")
```

## Calibration

Calibration assesses how well predicted probabilities match observed outcomes.

```{r calibration-metrics}
if (!is.null(rd$calibration$slope)) {
  cal_metrics <- data.frame(
    Metric = c("Calibration Slope", "Calibration Intercept", "ECE"),
    Value = c(
      sprintf("%.3f", rd$calibration$slope),
      sprintf("%.3f", rd$calibration$intercept %||% NA),
      sprintf("%.3f", rd$calibration$ece %||% NA)
    )
  )
  knitr::kable(cal_metrics, caption = "Calibration Metrics")
} else {
  cat("Calibration metrics not computed for this analysis.")
}
```

## Feature Stability

Feature stability measures how consistently features are selected across cross-validation folds. High stability indicates robust signal; low stability suggests potential overfitting.

```{r stability-summary}
if (!is.null(rd$stability$nogueira_index)) {
  cat(sprintf("**Nogueira Stability Index:** %.4f\n\n", rd$stability$nogueira_index))
  cat(sprintf("**Interpretation:** %s\n\n", rd$stability$interpretation %||% ""))
  cat(sprintf("**Consensus Features (selected in all folds):** %d\n",
              length(rd$stability$consensus_features %||% character(0))))
} else {
  cat("Stability analysis not available.")
}
```

```{r stability-heatmap, eval=!is.null(rd$stability$selection_frequency), fig.height=6}
if (!is.null(rd$stability$selection_frequency)) {
  freq <- rd$stability$selection_frequency
  top_features <- names(sort(freq, decreasing = TRUE))[1:min(20, length(freq))]

  freq_df <- data.frame(
    Feature = factor(top_features, levels = rev(top_features)),
    Frequency = freq[top_features]
  )

  ggplot(freq_df, aes(x = Frequency, y = Feature)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_vline(xintercept = 1.0, linetype = "dashed", color = "red") +
    scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
    labs(
      title = "Top 20 Features by Selection Frequency",
      x = "Selection Frequency Across Folds",
      y = NULL,
      caption = "Red line indicates 100% selection (consensus features)"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
}
```

---

# Reproducibility

## Software Environment

```{r session-info}
cat(sprintf("OmicSelector Version: %s\n", rd$meta$omicselector_version %||% "N/A"))
cat(sprintf("R Version: %s\n", rd$meta$r_version %||% "N/A"))
if (!is.null(rd$reproducibility$seed)) {
  cat(sprintf("Random Seed: %d\n", rd$reproducibility$seed))
}
```

## Configuration

```{r config, eval=!is.null(rd$meta$config)}
if (!is.null(rd$meta$config)) {
  cat("```yaml\n")
  cat(yaml::as.yaml(rd$meta$config))
  cat("```\n")
}
```

---

# TRIPOD+AI Compliance Checklist

| Item | Description | Status |
|------|-------------|--------|
| 1 | Title identifies as prediction model study | ✓ |
| 4 | Data source described | `r ifelse(!is.null(rd$data$n_samples), "✓", "○")` |
| 5 | Validation approach described | ✓ Nested CV |
| 6 | Outcome defined | `r ifelse(!is.null(rd$data$target_name), "✓", "○")` |
| 7 | Predictors defined | `r ifelse(!is.null(rd$model$filter_method), "✓", "○")` |
| 10 | Statistical analysis described | ✓ |
| 15 | Model performance reported | `r ifelse(!is.null(rd$discrimination$auc), "✓", "○")` |
| 16 | Calibration assessed | `r ifelse(!is.null(rd$calibration$slope), "✓", "○")` |

**Legend:** ✓ = Addressed, ○ = Partially addressed or N/A

---

*Report generated by OmicSelector 2.0 - Zero-Leakage Biomarker Discovery*

*"Optimization without validation is hallucination."*
